---
title: "Example Rmarkdown document: estimating pi"
author: Stephen J Eglen
bibliography: refs.bib
output:
  bookdown::html_document2: default
  bookdown::pdf_document2:
    keep_tex: true
  bookdown::word_document2:
    toc: true
---


## Disclaimer

Not everything is yet working (October 2022) in this document,
compared to the Rnw version.  Compared to LaTeX, I think there is less
flexibility for writing papers/reports, but for simple documents, the
Rmd approach may be quicker to write and sufficient for your needs.
Bookdown is a system for writing books, but can also be used for
writing reports, as is done here.


# Introduction

This is an example document created using the rmarkdown/bookdown system
<https://bookdown.org/>.  Rmarkdown is a tool for combining both
documentation and R code within the same file, similar to Sweave.  For
this document, the master file is `estimatek.Rmd`.  This is processed
by rmarkdown in R, which runs the R code to generate textual/graphical
output, and also creates a HTML document.  On recent machines, once
rmarkdown and kntir are installed, you should be able to generate the
pdf using:

```{r eval=FALSE,include=TRUE}
require(rmarkdown)
render('estimatek.Rnw')
```

Within RStudio, there is a handy "Knit HTML"  button.

Both estimatek.Rmd and estimatek.html are available from:
https://github.com/lgatto/spr/tree/master/estimate (you will need
`estimatek.Rmd` and `diff.R`)

Knitr and rmarkdown have built in support for caching.  By default it is
turned off in this document, but change FALSE to TRUE in the following
code-chunk and see if you can work out what happens:

```{r setup}
require(xtable)
require(printr)
require(knitr) ## for kable
knitr::opts_chunk$set(cache=FALSE)
```

#  Task: estimate the value of $\pi$

Our task is to estimate the value of $\pi$ by simulating darts being
thrown at a dartboard.  Imagine that the person throwing the darts is
not very good, and randomly throws each dart so that it falls
uniformly within a square of side length $2r$, with the dartboard of
radius $r$ centred within that square.  If the player throws $n$
darts, and $d$ of them hit the dartboard, then for large enough $n$,
the ratio $d/n$ should approximate the ratio of the area of the
dartboard to the enclosing square, $\pi r^2 / 4 r^2 \equiv \pi/4$.
From this, we can estimate $\pi \approx 4d/n$.

We start with an example, see Figure \@ref(fig:dartboard), using R to
draw both the dartboard and the surrounding square, together with
`n=50` darts.  The radius of the dartboard here is 1 unit, although
the value is not important.

```{r dartboard, fig.cap='The dartboard method for estimating pi.'}
r <- 1
n <- 50
par(las=1)
plot(NA, xlim=c(-r,r), ylim=c(-r,r), asp=1, bty='n',
     xaxt='n', yaxt='n', xlab='', ylab='')
axis(1, at=c(-r,0,r)); axis(2, at=c(-r,0,r))
symbols(x=0, y=0, circles=r, inch=F, add=T)
x <- runif(n, -r, r);  y <- runif(n, -r, r)
inside <- (x^2 + y^2) < r^2
d <- sum(inside)
points(x, y, pch=ifelse(inside, 19, 4))
rect(-r, -r, r, r, border='blue', lwd=2)
```

A dart is drawn as a filled circle if it falls within the dartboard,
else it is drawn as a cross.  In this case the number of darts within
the circle is `r d`, and so the estimated value is $\pi \approx
`r {4*d/n}`$.

The estimate of $\pi$ should improve as we increase the number of darts
thrown at the dartboard.  To verify this, we write a short function
that, given the number of darts to throw, $n$, returns an estimate of
$\pi$.

```{r}
estimate.pi <- function(n=1000) {
  ## Return an estimate of PI using dartboard 
  ## method with N trials.
  r <- 1 ## radius of dartboard
  x <- runif(n, min=-r, max=r)
  y <- runif(n, min=-r, max=r)
  l <- sqrt(x^2 + y^2)
  d <- sum(l<r)
  4*d/n
}
```

We can then test the procedure a few times, using the default number
of darts, 1000:

```{r}
replicate(9, estimate.pi())
```
Finally, for a given value of $n$, we can show 99 estimates of $\pi$,
as clearly the estimate will vary from run to run.  In
Figure \@ref(fig:n), we compare the estimates of $\pi$ for three
different values of $n$.


```{r n, fig.cap='Estimates of Pi as we increase number of darts' }
ns <- 10^c(2,3,4)
res <- lapply(ns, function(n) replicate(99, estimate.pi(n)))
par(las=1, bty='n')
stripchart(res, method="jitter", group.names=ns, 
           xlab="number of darts",
           ylab=expression(paste('estimate of ', pi)),
           vert=TRUE, pch=20, cex=0.5)
abline(h=pi, col='red')
```

<!-- 
\caption{Mone-Carlo estimates of $\pi$ improve as n increases.  Red line
denotes the true value of $\pi$.}
\label{fig:n}
-->
 
# printr

The printr package provides a convenient interface for making tables.
Here's a simple example.

```{r}
df = data.frame(name=c("joe", "ann", "bob"),
  age=c(19, 24, 27),
  height=c(1.8, 1.75, 1.7))
df
```

Or you can add a caption, see Table \@ref(tab:kable) below.




```{r kable}
kable(df, caption="Example output from kable")
```


# Listing

Finally, an example listing.  We read in all the code from a given
file (here diff.R) and then assign it to the ``all'' label which we
then include without evaluation, or showing it.  We then can show the
code in the Appendix of this document.  You could use this as a way of
putting your code in an appendix, but using it in the main part of the
document.


```{r external-code,chunk=FALSE,include=FALSE,results='hide'}
read_chunk('diff.R',labels='all')
```

```{r all,echo=FALSE,results='hide'}
```

```{r}
x <- c(5.9, 10.2, 12.4, 18.8)
all.equal(diff1(x), diff2(x))
```


See http://yihui.name/knitr/demo/externalization/ for further
information.

# References

References can be included from your bib file: [@R-base]  [@Xie2016-he].

## Compiling this document

```{r eval=FALSE}
rmarkdown::render('estimatek.Rmd')
```

# Appendix

```{r all,eval=FALSE}
```

# References

